---
title: 'Co_2 1997'
output: 'pdf_document'  
---

```{r load packages, echo = FALSE, message = FALSE}
library(tidyverse)
library(ggplot2)
library(feasts)
library(tsibble)
library(latex2exp)
library(patchwork)
library(fable)
library(forecast)
library(stargazer)

theme_set(theme_minimal())
knitr::opts_chunk$set(dpi=1000)
```

```{r, echo = FALSE, message = FALSE}
co2 <- as_tsibble(co2) %>% filter(year(index)<1998)

```

# Data

The data measures the monthly average atmospheric CO2 concentration from 1959 to 1997, expressed in parts per million (ppm). It was initially collected by a infrared gas analyzer installed at Mauna Loa in Hawaii, which was one of the four analyzers installed by Keeling to evaluate whether there was a persistent increase in CO2 concentration. 

Fig.1 shows a clear long-term upward trend, which is confirmed by Fig.2 where the growth rate for each year is above zero. Fig.2 also suggests the average growth rate after 1970 is higher than that before 1970, although there's no evidence of accelerating growth.

Another feature of the data is its robust seasonal pattern, with peak in May and bottom in October almost every year (see Fig.3). Keeling believes it was the result of the activity of land plants.

Fig.4 is the histogram of the remaining or irregular components after removing the trend and the seasonal components from the data with STL^[Cleveland, R. B., Cleveland, W. S., McRae, J. E., & Terpenning, I. J. (1990). STL: A seasonal-trend decomposition procedure based on loess. Journal of Official Statistics, 6(1), 3–33.]. It looks like a normal distribution without obvious outliers.  


```{r, echo = FALSE, message = FALSE, warning=FALSE,fig.height=3}
p1 <- autoplot(co2) +
  ggtitle("Fig.1 Atmospheric CO2 concentration\n monthly average, parts per million (ppm) ") +
  xlab(NULL) + ylab(NULL)+ 
  theme(text = element_text(size = 8)) 
p2 <- co2 %>% index_by(year = year(index)) %>%
  summarise(annual_avg = mean(value)) %>%
  mutate(annual_growth = (annual_avg / lag(annual_avg, 1) - 1) * 100) %>%
  autoplot(.vars = annual_growth) +
  xlab(NULL) + ylab(NULL) +
  ggtitle("Fig.2 Annual growth rate of concentration, %")+ 
  theme(text = element_text(size = 8)) 
p3 <- gg_season(co2) +
  xlab(NULL) + ylab(NULL) +
  ggtitle("Fig.3 Seasonal plot of CO2 concentration")+ 
  theme(text = element_text(size = 8)) 
p4 <- co2 %>% model(STL(value ~ trend(window = 120) + season(window = "periodic"),
                        robust = TRUE)) %>%
  components() %>% pull(remainder) %>% gghistogram() +
  ggtitle("Fig.4 Histogram of irregular\n component by STL")+ 
  theme(text = element_text(size = 8)) 
# p3 <- ggAcf(co2$value)
# p4 <- ggPacf(co2$value)
(p1 | p2) / (p3 | p4)
```

# Linear model

Since the long term trend of the $CO_2$ data looks linear and the variation around the trend seems stable, a log transformation of the data is not necessary (which is also supported by the stable and symmetric residuals in Fig.5) and we can first fit the original data with a linear time trend model as:

\begin{equation}
\label{eq:one}
\text{CO}_{2} = \beta_0 + \beta_1t + \epsilon_{t},
\end{equation} 

which gives the parameters as:

\begin{equation}
\label{eq:two}
\text{CO}_{2} = 311.5 + 0.11t + \epsilon_{t}
\end{equation}

This linear trend model implies that the $CO_2$ concentration increased 0.11/month on average during 1959 to 1997. However, the residuals plots in Fig.5 suggest this simple linear trend model is not adequate in the following two aspects. 

First, the mean of the residual forms a "U" shape along time, suggesting a quadratic or higher order polynomial time trend model may be more appropriate. For instance, the residual from a quadratic time trend model shows a more constant mean over time, as shown in Fig.6.

```{r, echo = FALSE, message = FALSE, warning=FALSE}

fit <-co2 %>% model(
  linear_trend =TSLM(value~trend()),
  quadratic_trend=TSLM(value~trend()+I(trend()^2)))

fit %>% select(linear_trend) %>% gg_tsresiduals() + ggtitle("Fig.5 Residual plot of the linear trend model")
fit %>% select(quadratic_trend) %>% gg_tsresiduals() + ggtitle("Fig.6 Residual plot of the quadratic trend model")

```

In addition, the ACF plot in Fig.5 indicates strong seasonal patterns exists in the residuals, suggesting we should consider seasonal factors in the model, and one solution is to include 11 dummy variables in the model to indicate the 12 months.

Based on the two points above, we compare the following 3 candidate models:
\begin{equation}
\label{eq:three}
\text{Linear time trend: }CO_{2} = \alpha + \beta_0t +\gamma_1D_{Jan}+\gamma_2D_{Feb}+...+\gamma_{11}D_{Nov} + \epsilon_{t}
\end{equation} 
\begin{equation}
\label{eq:four}
\text{Quadratic time trend: CO}_{2} = \alpha + \beta_0t + \beta_1t^2 +\gamma_1D_{Jan}+\gamma_2D_{Feb}+...+\gamma_{11}D_{Nov} + \epsilon_{t}
\end{equation} 
\begin{equation}
\label{eq:five}
\text{Cubic time trend: CO}_{2} = \alpha + \beta_0t + \beta_1t^2 + \beta_2t^3 + \gamma_1D_{Jan}+\gamma_2D_{Feb}+...+\gamma_{11}D_{Nov} + \epsilon_{t}
\end{equation} 

We use the data before 1991 as the training set and the rest as the validation set (similar to a 80-20 split). The final choice of the model depends on the combination of 3 guidelines: 1) whether the coefficients of the model make sense; 2) the root mean square error (RMSE) on the validation set; 3) the AICc on the training set. The details of model fitting results and validation performance are included in the Appendix. Although both RMSE and AICc favor the cubic model, its negative cubic term indicates that after a certain period the $CO_2$ concentration will start to decrease automatically, which makes little sense. For the rest two, although the quadratic model has better AICc on the training set, it has a higher RMSE on the validation set, which may indicate an overfiting. Therefore, the linear time trend model becomes our final choice.

```{r, echo = FALSE, message = FALSE, warning=FALSE}
co2_copy <- co2 %>% append_row(276) %>%
  mutate(
    num_index = time(index),
    num_index_qudratic = num_index ^ 2,
    num_index_cubic = num_index ^ 3,
  )
for (i in 1:11) {
  name =
    co2_copy <-
    co2_copy %>% mutate("month_{i}" := ifelse(month(index) == i, 1, 0))
}
co2_training = co2_copy %>% filter(year(index) < 1991)
co2_valid = co2_copy %>% filter(year(index) < 1998, year(index) >= 1991)
co2_forecast = co2_copy %>% filter(year(index) >= 1998)

dummy_name = paste0("month_", 1:11, collapse = "+")
fit <- co2_training |>
  model(
    model_linear = TSLM(as.formula(paste0(
      "value ~ num_index   +", dummy_name
    ))),
    model_quadratic = TSLM(as.formula(
      paste0("value ~ num_index  + num_index_qudratic +", dummy_name)
    )),
    model_cubic = TSLM(as.formula(
      paste0(
        "value ~ num_index + num_index_qudratic + num_index_cubic +",
        dummy_name
      )
    ))
  )

vd <- forecast(fit, new_data = co2_valid)
# co2_training %>% filter(index >= yearmonth("1985M01")) %>% autoplot(value, PI = FALSE) +
#   autolayer(vd) + autolayer(co2_valid)

# although the result above prefer the cubic model, I would suggest a linear or quadratic one.
final_model_poly <- co2_copy %>% filter(year(index) < 1998) %>%
  model(TSLM(as.formula(paste0(
    "value ~ num_index + ", dummy_name
  ))))
fc_linear <- final_model_poly %>% forecast(co2_forecast)
co2_copy %>% filter(year(index) < 1998) %>% autoplot(value) + autolayer(fc_linear)
```


# The rest should go to Appendix

```{r}
print("AIC etc on training set")
glance(fit) |> arrange(AICc) |> select(.model,AIC,AICc,BIC)

print("RMSE on validation set")

fabletools::accuracy(vd,co2_valid)|> select(.model,RMSE)

report(fit%>% select(model_cubic))
```

