---
title: 'Co_2 1997'
output: 'pdf_document'  
---

```{r load packages, echo = FALSE, message = FALSE}
library(tidyverse)
library(ggplot2)
library(feasts)
library(tsibble)
library(latex2exp)
library(patchwork)
library(fable)
library(forecast)
library(stargazer)
library(dplyr)
library(knitr)
theme_set(theme_minimal())
knitr::opts_chunk$set(dpi=1000)
```

```{r, echo = FALSE, message = FALSE}
co2 <- as_tsibble(co2) %>% filter(year(index)<1998)

```

# Data

The data measures the monthly average atmospheric CO2 concentration from 1959 to 1997, expressed in parts per million (ppm). It was initially collected by a infrared gas analyzer installed at Mauna Loa in Hawaii, which was one of the four analyzers installed by Keeling to evaluate whether there was a persistent increase in CO2 concentration. 

Fig.1 shows a clear long-term upward trend, which is confirmed by Fig.2 where the growth rate for each year is above zero. Fig.2 also suggests the average growth rate after 1970 is higher than that before 1970, although there's no evidence of accelerating growth.

Another feature of the data is its robust seasonal pattern, with peak in May and bottom in October almost every year (see Fig.3). Keeling believes it was the result of the activity of land plants.

Fig.4 is the histogram of the remaining or irregular components after removing the trend and the seasonal components from the data with STL^[Cleveland, R. B., Cleveland, W. S., McRae, J. E., & Terpenning, I. J. (1990). STL: A seasonal-trend decomposition procedure based on loess. Journal of Official Statistics, 6(1), 3–33.]. It looks like a normal distribution without obvious outliers.  


```{r, echo = FALSE, message = FALSE, warning=FALSE,fig.height=5}
p1 <- autoplot(co2) +
  ggtitle("Fig.1 Atmospheric CO2 concentration\n monthly average, parts per million (ppm) ") +
  xlab(NULL) + ylab(NULL)+ 
  theme(text = element_text(size = 8)) 
p2 <- co2 %>% index_by(year = year(index)) %>%
  summarise(annual_avg = mean(value)) %>%
  mutate(annual_growth = (annual_avg / lag(annual_avg, 1) - 1) * 100) %>%
  autoplot(.vars = annual_growth) +
  xlab(NULL) + ylab(NULL) +
  ggtitle("Fig.2 Annual growth rate of concentration, %")+ 
  theme(text = element_text(size = 8)) 
p3 <- gg_season(co2) +
  xlab(NULL) + ylab(NULL) +
  ggtitle("Fig.3 Seasonal plot of CO2 concentration")+ 
  theme(text = element_text(size = 8)) 
p4 <- co2 %>% model(STL(value ~ trend(window = 120) + season(window = "periodic"),
                        robust = TRUE)) %>%
  components() %>% pull(remainder) %>% gghistogram() +
  ggtitle("Fig.4 Histogram of irregular\n component by STL")+ 
  theme(text = element_text(size = 8)) 
# p3 <- ggAcf(co2$value)
# p4 <- ggPacf(co2$value)
(p1 | p2) / (p3 | p4)
```

# Linear model

Since the long term trend of the $CO_2$ data looks linear and the variation around the trend seems stable, a log transformation of the data is not necessary (which is also supported by the stable and symmetric residuals in Fig.5) and we can first fit the original data with a linear time trend model as:

\begin{equation}
\label{eq:one}
\text{CO}_{2} = \beta_0 + \beta_1t + \epsilon_{t},
\end{equation} 

which gives the parameters as:

\begin{equation}
\label{eq:two}
\text{CO}_{2} = 311.5 + 0.11t + \epsilon_{t}
\end{equation}

This linear trend model implies that the $CO_2$ concentration increased 0.11/month on average during 1959 to 1997. However, the residuals plots in Fig.5 to Fig.7 suggest this simple linear trend model is not adequate in the following two aspects. 

First, the mean of the residual forms a "U" shape along time, suggesting a quadratic or higher order polynomial time trend model may be more appropriate. For instance, the residual from a quadratic time trend model shows a more constant mean over time, as shown in Fig.8.

```{r, echo = FALSE, message = FALSE, warning=FALSE,fig.height=3}
fit <- co2 %>% model(
  linear_trend = TSLM(value ~ trend()),
  quadratic_trend = TSLM(value ~ trend() + I(trend() ^ 2))
)

resid_linear <-
  fit %>% dplyr::select(linear_trend) %>% residuals()
resid_quadratic <-
  fit %>% dplyr::select(quadratic_trend) %>% residuals()
p5 <-
  autoplot(resid_linear) + ggtitle("Fig.5 Residual of the linear trend model") +
  theme(text = element_text(size = 8))
p6 <-
  ggAcf(resid_linear) + ggtitle("Fig.6 ACF of the linear trend model residuals") +
  theme(text = element_text(size = 8))
p7 <-
  gghistogram(resid_linear %>% pull(.resid)) + ggtitle("Fig.7 Histogram of the linear\n trend model residuals") +
  theme(text = element_text(size = 8))
p8 <-
  autoplot(resid_quadratic) + ggtitle("Fig.8 Residual of the quadratic\n time trend model") +
  theme(text = element_text(size = 8))
(p5 | p6) / (p7 | p8)
```

In addition, Fig.6, the ACF plot, indicates strong seasonal patterns exists in the residuals, suggesting we should consider seasonal factors in the model, and one solution is to include 11 dummy variables in the model to indicate the 12 months.

Based on the two points above, we compare the 2 candidates: a quadratic time trend model and a cubic one, as below.

\begin{equation}
\label{eq:three}
\text{Quadratic time trend: CO}_{2} = \alpha + \beta_0t + \beta_1t^2 +\gamma_1D_{Jan}+\gamma_2D_{Feb}+...+\gamma_{11}D_{Nov} + \epsilon_{t}
\end{equation} 
\begin{equation}
\label{eq:four}
\text{Cubic time trend: CO}_{2} = \alpha + \beta_0t + \beta_1t^2 + \beta_2t^3 + \gamma_1D_{Jan}+\gamma_2D_{Feb}+...+\gamma_{11}D_{Nov} + \epsilon_{t}
\end{equation} 

We use the data before 1991 as the training set and the rest as the validation set (similar to a 80-20 split). Our final choice of the model depends on the combination of 2 guidelines: 1) the information criterion (AICc, BIC) from the model fitting process and 2) the root mean square error (RMSE) of predictions on the validation set, which are listed in Table.1. Both information criterion (AICc, BIC) and RMSE favor the cubic model. Therefore, the cubic time trend model becomes our final choice. Its details are in the Appendix. We plot the forecast of this model till 2020 in Fig.7. One thing to note is that because the coefficient of the cubic term is negative, the predicted values will eventually begin to decrease when predicting the far future. In fact we can see from Fig.7 that the predicted values have almost topped. If it doesn't make sense, we should confine our predicting interval to the near term.


```{r, echo = FALSE, message = FALSE, warning=FALSE,fig.height=2}
co2_training = co2 %>% filter(year(index) < 1991)
co2_valid = co2 %>% filter(year(index) < 1998, year(index) >= 1991)
fit_poly <- co2_training |>
  model(
    quadratic = TSLM(value ~ trend() + I(trend() ^ 2) + season()),
    cubic = TSLM(value ~ trend() + I(trend() ^ 2) + I(trend() ^ 3) +
                   season())
  )

vd <- forecast(fit_poly, h = 72)
fc_poly <- co2 %>%
  model(TSLM(value ~ trend() + I(trend() ^ 2) + I(trend() ^ 3) +
               season())) %>%
  forecast(h = 276)
model_ic = glance(fit_poly) %>%  dplyr::select(.model, AIC, AICc, BIC) %>% arrange(AICc)
vc_acc = fabletools::accuracy(vd, co2_valid) |> dplyr::select(.model, RMSE)
compare = cbind(model_ic, vc_acc['RMSE'])
kable(compare, caption = "Information Criterion of model fitting and RMSE of validation")
co2 %>%  autoplot(value) + autolayer(fc_poly) + ggtitle("Fig.7 Forecasts of CO2 level Up To 2020 Using Polynomial Trend Time Model") +
  theme(text = element_text(size = 8)) 
```


# The rest should go to Appendix

```{r}
fit_poly %>% dplyr::select(cubic) %>% report()
```


